---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CCCtools

<!-- badges: start -->
<!-- badges: end -->

The goal of CCCtools is to provide functions to run [CellChat v2](https://github.com/jinworks/CellChat) and [CellPhoneDB v5](https://github.com/ventolab/CellphoneDB/tree/master) as well as compare their outputs. Gokce was here!

## Installation

You can install the development version of CCCtools:

```{r, eval=FALSE}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load_gh("nigelhojinker/CCCtools")
```


## Example data

`seu.NL` is the Seurat object containing 2,233 cells from non-lesional human skin from four different patients. This dataset was created from the data used originally in CellChat tutorial. See this [script for how the data was generated](data-raw/demo_data.md).

```{r example}
suppressPackageStartupMessages(library(CCCtools)) 

data(seu.NL)

seu.NL
```

## Tutorials

In **CCCtools**, we provide the functions `run_cellchat()` and `run_cellphonedb()` for users to perform CellChat and CellPhoneDB (method 2) analysis directly on their processed Seurat object. 

Please refer to the following links to the documentations on running these tools on your dataset:

- [Running CellChat on my Seurat object](../data-raw/Run_CellChat.md)
- [Running CellPhoneDB on my Seurat object](../data-raw/Run_CellPhoneDB.md)
