---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CCCtools

<!-- badges: start -->
<!-- badges: end -->

The goal of CCCtools is to provide functions to run [CellChat v2](https://github.com/jinworks/CellChat) and [CellPhoneDB v5](https://github.com/ventolab/CellphoneDB/tree/master) as well as compare their outputs. Gokce was here!

## Installation

You can install the development version of CCCtools:

```{r, eval=FALSE}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load_gh("nigelhojinker/CCCtools")
```


## Example data

`seu.NL` is the Seurat object containing 2,233 cells from non-lesional human skin from four different patients. This dataset was created from the data used originally in CellChat tutorial. See this [script for how the data was generated](data-raw/demo_data.md).

```{r example}
suppressPackageStartupMessages(library(CCCtools)) 

data(seu.NL)

seu.NL
```


